# Preparing the workspace
* Preparations: clear the workspace en set the working directory 
* make sure that the rounding of integers in the way desired: broman packages gives access to the function 'myround'
* make sure that lattice is loaded so xyplot can be used for multipanel-presenatation of data  
```{r prepare the workspace, read and preprocess the data}
rm(list=ls())
setwd("~/___Big Data/Coursera/05 Reproducible Research/RepData_PeerAssessment1")
library(broman)
library(lattice)
```
# Loading and Preprocessing the data
* read the data into a dataframe called "dataset"  
* make sure the date in the dataset is really a Date-object
* give an overview of the dataset to check type and format: str and head  

```{r loading and preprocessing the data}
dataset <- read.csv("activity.csv", stringsAsFactors=FALSE, header = T, sep = ",")
dataset$date = as.Date(dataset$date, "%Y-%m-%d")
str(dataset)
head(dataset)
```
#What is mean total number of steps taken per day?
* sum steps per day (all the intervals on the same date) - use aggregate for that
* give insight in the df that was created with head and str
* make a histogram of the totals
* then calculate the mean per day

```{r number of steps per day}
stepsPerDay = aggregate(dataset$steps, by = list(dataset$date), FUN=sum, na.rm=TRUE)
names(stepsPerDay) <- c("date", "total")
str(stepsPerDay)
head(stepsPerDay)
hist(stepsPerDay$total,
     breaks=seq(from=0, to=25000, by=1250), 
     col="blue", 
     xlab="Total number of steps", 
     ylab="Frequency", 
     main="61 days:frequency of numbers of steps")
meanSteps = mean(stepsPerDay$total)
medianSteps = median(stepsPerDay$total)
```

The mean number of steps per day is `r myround(meanSteps,2)`.   
The median number of steps per days is `r myround(medianSteps,2)`.  

#What is the average daily activity pattern?
* sum steps by interval - use aggregate again, give the variables the names 'interval' and 'steps', and give the first six lines
* make a plot of the activity per interval over the entire time period
* then calculate max number of steps and the interval that goes with it.

```{r daily pattern}
stepsPerInterval = aggregate(dataset$steps, by = list(dataset$interval), FUN=mean, na.rm=TRUE)
names(stepsPerInterval) <- c("interval", "steps")
head(stepsPerInterval)
plot(stepsPerInterval$interval,stepsPerInterval$steps,type = "l", col="black",  xlab="Intervals", ylab="Number of steps", main="Number of steps per interval")
maxSteps = round(max(stepsPerInterval$steps),0)
mostActiveInterval <- stepsPerInterval[stepsPerInterval$steps==max(stepsPerInterval$steps),1]
```

Averagely this subject is most active around interval number `r mostActiveInterval`. He or she takes on average `r maxSteps` steps during that interval.

#Imputing missing values  
##how many missing values are there?
* Find out the missing values by applying is.na to the steps-field in the dataframe 
* if there is a missing value it will return TRUE -- which counts as 1  
* if not it will return FALSE -- which counts as 0
* add them up and you will have the total of the missing values  
```{r how many NA}
NA_count <- sum(is.na(dataset$steps))
```
There are `r NA_count` missing values in the steps-field of the dataset.

##replace them by the mean of the interval of the missing value 
* an NA in interval 5 on particular day will be replaced by the mean of interval 5 in general 
* therefore:  
* find the positions that have an NA in steps: posNA  
* create a new dataset to preserve the original set: nwdataset  
* and replace the NA's on posNA with mean values from the stepPerInterval-df  
* give an overview of nwdataset with str and head  
```{r replace NA with mean value of that interval}
posNA <- which(is.na(dataset$steps))
nwdataset <- dataset
for (i in 1:length(posNA)){
        nwdataset[posNA[i],1] <- stepsPerInterval[stepsPerInterval$interval ==                   dataset[posNA[i],3],2]
}
str(nwdataset)
head(nwdataset)
```

##make a histogram of the total number of steps in nwdataset
* make a histogram with the new values  
* calculate and report the mean and median total number of steps taken per day
* make sure the difference between the first and second calculation is obvious
* interpret the difference

```{r histogram with NA replaced by mean of that interval}
nwStepsPerDay = aggregate(nwdataset$steps, by = list(nwdataset$date), FUN=sum)
names(nwStepsPerDay) <- c("date", "total")
hist(nwStepsPerDay$total, breaks=seq(from=0, to=25000, by=1250), col="black",     xlab="Total number of steps", ylab="Frequency", main="Number of steps per day")
nwMeanSteps = mean(nwStepsPerDay$total)
nwMedianSteps = median(nwStepsPerDay$total)
```

The mean number of steps per day was `r myround(meanSteps,2)` but with the NA's replaced by the mean it is `r myround(nwMeanSteps,2)`.  
The median number of steps per day was `r myround(medianSteps,2)` but with the NA's replaced by the mean it is `r myround(nwMedianSteps,2)`.
Imputing of NAs has the following effect: Both the mean and the median number of steps have increased. There are considerable less days with 0 steps (2 versus 10) and considerable more days with 10000 tot 11250 steps.  

#Are there differences in activity patterns between weekdays and weekends?  

##figure out which dates are weekend and which are workdays
* set the time to English (or the weekdays will be named in Dutch)
* create a new variable: weekday (by applying the function weekdays to the Date object) 
* and bind that variable to the dataset
* create a new variable: daytype by determining whether the day is a weekendday or not (ifelse)
* and bind that variable to the dataset as well
* give an overview of the amended dataset: str en head

```{r workdays vs weekends}
Sys.setlocale("LC_TIME", "English")
nwdataset <-cbind(nwdataset,day = weekdays(nwdataset$date, abbreviate = FALSE))
nwdataset <- cbind(nwdataset, daytype = ifelse(nwdataset$day == "Saturday" | nwdataset$day == "Sunday", "weekend", "weekday"))
str(nwdataset)
head(nwdataset)
stepsByDay <- aggregate(steps ~ interval + daytype, data = nwdataset, mean)
names(stepsByDay) <- c("interval", "daytype", "steps")
str(stepsByDay)
head(stepsByDay)
```
  
##draw a plot using lattice  
```{r}
xyplot(steps ~ interval | daytype, stepsByDay, type = "l", layout = c(1, 2), 
    xlab = "Interval", ylab = "Number of steps")
```
  
On weekdays there is one peak. During weekdays there is not so much of a peak but a higher activity overall.



